Classless Java

### 2016-03-24 OOPSLA16/sections/CaseStudy.tex
Spellcheck.

\begin{figure*}
\centering
\saveSpaceFig
\begin{tabular}{l|l}
\lstinputlisting[linerange=33-43]{../UseMixinLombok/src/casestudy/ep/TestExpression.java}% APPLY:linerange=EXPRESSION_INIT
&
\lstinputlisting[linerange=54-64]{../UseMixinLombok/src/casestudy/ep/TestExpression.java}% APPLY:linerange=EXPRESSION_PRINT
\end{tabular}
\caption{The Expression Problem (left: initial system, right: code for adding
  print operation).}\label{fig:ep}
\end{figure*}% \bruno{Yanlin: code is overflowing the margins. Note that
  % there are 2 spare lines that can be used on the right!}

@@

The \mixin annotation can also be used to create fluent interfaces.  When
creating fluent interfaces with \mixin, there are two main advantages:
\begin{enumerate}
\item Instead of forcing programmers to hand write code using \textbf{return
    this}, our approach with \mixin annotation removes this verbosity and
  automatically generates fluent setters.
\item The approach supports extensibility: the return types of fluent setters are
  automatically refined.
\end{enumerate}

@@

\paragraph{Extending the Query Language} The previous query language can be extended with a new feature
\texttt{orderBy} which orders the result records by a field that users
specify. With \mixin programmers just need to extend the interface \texttt{Database} with new
features, and the return type of fluent setters in
\texttt{Database} is automatically refined to \texttt{ExtendedDatabase}:
\lstinputlisting[linerange=14-18]{../UseMixinLombok/src/casestudy/sql/DatabaseTest.java}% APPLY:linerange=FLUENT_DATABASE_EXT
In this way, when a query created using \texttt{ExtendedDatabase},
all the fluent setters return the correct type, and not the old \texttt{Database} type, which would prevent calling
\texttt{orderBy}.

@@

\subsection{Refactoring an Interpreter}
The last case study applies our approach to a slightly larger code base of an
interpreter for a Lisp-like language
\lstinline{Mumbler}\footnote{https://github.com/cesquivias/mumbler/tree/master/simple},
which is created for tutorial purpose of Truffle~\cite{wurthinger2013one}. Keeping a balance between
simplicity and useful features, \lstinline{Mumbler} contains numbers, booleans,
lists (encoding function calls and special forms such as if-expression, lambdas,
etc). In the original code base, only one operation \texttt{eval} is
supported. We add one more operation \texttt{print} to the code base. Then we
modularize the code using our techniques to convert as many classes into
interfaces, separate out \texttt{eval} and \texttt{print} operation then combine
them. Although the code size is slightly increased (by 2.2\% SLOC), the
modularity is greatly increased.

>>>>

\begin{figure*}
\centering
\saveSpaceFig
\begin{tabular}{l|l}
\lstinputlisting[linerange=33-43]{../UseMixinLombok/src/casestudy/ep/TestExpression.java}% APPLY:linerange=EXPRESSION_INIT
&
\lstinputlisting[linerange=54-64]{../UseMixinLombok/src/casestudy/ep/TestExpression.java}% APPLY:linerange=EXPRESSION_PRINT
\end{tabular}
\caption{The Expression Problem (left: initial system, right: code for adding
  print operation).}\label{fig:ep}
\saveSpaceFig
\end{figure*}% \bruno{Yanlin: code is overflowing the margins. Note that
  % there are 2 spare lines that can be used on the right!}

@@

The \mixin annotation can also be used to create fluent interfaces.  When
creating fluent interfaces with \mixin, there are two main advantages:
\begin{enumerate}
\item Instead of forcing programmers to hand-write code using \textbf{return
    this}, our approach with \mixin annotation removes this verbosity and
  automatically generates fluent setters.
\item The approach supports extensibility: the return types of fluent setters are
  automatically refined.
\end{enumerate}

@@

\paragraph{Extending the Query Language} The previous query language can be extended with a new feature
\texttt{orderBy} which orders the result records by a field that users
specify. With \mixin programmers just need to extend the interface \texttt{Database} with new
features, and the return type of fluent setters in
\texttt{Database} is automatically refined to \texttt{ExtendedDatabase}:
\lstinputlisting[linerange=14-18]{../UseMixinLombok/src/casestudy/sql/DatabaseTest.java}% APPLY:linerange=FLUENT_DATABASE_EXT
In this way, when a query is created using \texttt{ExtendedDatabase},
all the fluent setters return the correct type, and not the old \texttt{Database} type, which would prevent calling
\texttt{orderBy}.

@@

\subsection{Refactoring an Interpreter}
The last case study applies our approach to a slightly larger code base of an
interpreter for a Lisp-like language
\lstinline{Mumbler}\footnote{https://github.com/cesquivias/mumbler/tree/master/simple},
which is created for tutorial purpose of Truffle~\cite{wurthinger2013one}. Keeping a balance between
simplicity and useful features, \lstinline{Mumbler} contains numbers, booleans,
lists (encoding function calls and special forms such as if-expression, lambdas,
etc). In the original code base, only one operation \texttt{eval} is
supported. We add one more operation \texttt{print} to the code base. Then we
modularize the code using our techniques to convert as many classes into
interfaces, separate out \texttt{eval} and \texttt{print} operations then combine
them. Although the code size is slightly increased (by 2.2\% SLOC), the
modularity is greatly increased.

### 2016-03-24 OOPSLA16/sections/FutureWork.tex
Spellcheck.

Using packages as a boundary we can make all object interfaces \emph{where some form of privateness is desirable}
as invisible types for the rest of the program.
This preserve encapsulation, but limits inheritance:
package protected interfaces can not be seen,
thus can not be implemented from outside the package.

@@

%\subsection{Clone, toString, equals and hashCode}
\paragraph{Clone, toString, equals and hashCode}
Methods originally defined in Java class \Q@Object@, as \Q@clone@ and
\Q@toString@, can be supported by our approach with special care. If an
interface annotated with \mixin asks an implementation for \Q@clone@,
\Q@toString@, \Q@equals@ or \Q@hashCode@ we can easily generate one from the
fields.\footnote{In particular, for clone we can do automatic return type
  refinement as we do for \Q@with-@ and fluent setters. Note how this would
  solve most of the Java ugliness related to \Q@clone@ methods.}  However, if
the user wishes to provide his own implementation, since the method is also
implemented in \Q@Object@, a conflict arises. The generated code can resolve the
conflict inside \Q@of@, by implementing the method and delegating it to the user
implementation, thus

@@

\paragraph{A real IB language}
In this article we presented the concept of IB as a programming pattern over Java.
An alternative would be to have a language providing IB as its only paradigm.
We do not wish to define here a complete IB language, just to show the reader IB as
a promising way to design a new oo language.
To define such language, the first task is to define the set of abstract state operations.
Of course we need getters; \Q@withX@ methods seams to be pretty convenient, and they enable a
clean function programming style that would be otherwise too cumbersome.

>>>>

Using packages as a boundary we can make all object interfaces \emph{where some form of privateness is desirable}
as invisible types for the rest of the program.
This preserves encapsulation, but limits inheritance:
package protected interfaces cannot be seen,
thus cannot be implemented from outside the package.

@@

%\subsection{Clone, toString, equals and hashCode}
\paragraph{Clone, toString, equals and hashCode}
Methods originally defined in Java class \Q@Object@, such as \Q@clone@ and
\Q@toString@, can be supported by our approach with special care. If an
interface annotated with \mixin asks an implementation for \Q@clone@,
\Q@toString@, \Q@equals@ or \Q@hashCode@ we can easily generate one from the
fields.\footnote{In particular, for clone we can do automatic return type
  refinement as we do for \Q@with-@ and fluent setters. Note how this would
  solve most of the Java ugliness related to \Q@clone@ methods.}  However, if
the user wishes to provide his own implementation, since the method is also
implemented in \Q@Object@, a conflict arises. The generated code can resolve the
conflict inside \Q@of@, by implementing the method and delegating it to the user
implementation, thus

@@

\paragraph{A real IB language}
In this article we presented the concept of IB as a programming pattern over Java.
An alternative would be to have a language providing IB as its only paradigm.
We do not wish to define here a complete IB language, just to show the reader IB as
a promising way to design a new OO language.
To define such language, the first task is to define the set of abstract state operations.
Of course we need getters; \Q@withX@ methods seams to be pretty convenient, and they enable a
clean function programming style that would be otherwise too cumbersome.

### 2016-03-24 OOPSLA16/sections/Implementation.tex
Spellcheck.

\subsection{Limitations}
Our prototype implementation has certain limitations:
\begin{itemize}
\item Eclipse has its own compiler (ejc), alternative to javac.
  Lombok allows writing handlers for either javac or ejc.
  At this stage our implementation only realizes the Eclipse handler and our
  experiments are all conducted in Eclipse. The implementation for
  the \texttt{javac} handler is still missing.
\item The current implementation supports generics in a simplified manner: 
type parameters can be used but the generic
method typing is neither formalized in this paper nor explicitly checked by our 
implementation, but simply  delegated to the Java compiler.
\item 
Lombok offers only limited/experimental support for separate compilation, that is
to access information of code defined in different files.
In the same spirit, we have a mature \mixin annotation, that
does not support separate compilation yet: it requires that all
  related interfaces have to appear in a single Java file.
Reusing the logic inside the yet experimental Lombok annotation \lstinline{@Delegate},
we also offer a less polished annotation supporting
separate compilation for files in the same package.
%  has the same behavior as \mixin. A limitation is
%  that users need to put references to the super types
% of the annotated type together with the annotation, for instance,
%  \lstinline{@Delegate(types = Point2D.class)}.
%   Another limitation is the methods generated by \lstinline{@Delegate} will
%  not be visible in the Outline window, but they can still be auto-completed in Eclipse.
\end{itemize}

>>>>

\subsection{Limitations}
Our prototype implementation has certain limitations:
\begin{itemize}
\item Eclipse has its own compiler (ejc), alternative to javac.
  Lombok allows writing handlers for either javac or ejc.
  At this stage our implementation only realizes the Eclipse handler and our
  experiments are all conducted in Eclipse. The implementation for
  the \texttt{javac} handler is still missing.
\item The current implementation supports generics in a simplified manner:
type parameters can be used but the generic
method typing is neither formalized in this paper nor explicitly checked by our
implementation, but simply  delegated to the Java compiler.
\item
Lombok offers only limited/experimental support for separate compilation, that is
to access information of code defined in different files.
In the same spirit, we have a mature \mixin annotation, which
does not support separate compilation yet: it requires that all
  related interfaces have to appear in a single Java file.
Reusing the logic inside the yet experimental Lombok annotation \lstinline{@Delegate},
we also offer a less polished annotation supporting
separate compilation for files in the same package.
%  has the same behavior as \mixin. A limitation is
%  that users need to put references to the super types
% of the annotated type together with the annotation, for instance,
%  \lstinline{@Delegate(types = Point2D.class)}.
%   Another limitation is the methods generated by \lstinline{@Delegate} will
%  not be visible in the Outline window, but they can still be auto-completed in Eclipse.
\end{itemize}

### 2016-03-24 OOPSLA16/sections/RelatedWork.tex
Spellcheck.

\paragraph{Models of Object-Oriented programming}
There are traditionally two big styles of OOP: class-based and
prototype-based. The class-based tradition started with Simula~\cite{Dahl66simula}, 
and is the dominant style of OOP today. Major languages such 
as Java, C\#, C++ or Scala favour a class based OOP style. Very often
class-based languages are statically typed and they have interfaces or
some other similar mechanisms such as traits. However, in contrast to IB
programming, interfaces or traits in CB programming are not meant to
replace classes, but rather, they complement classes. Normally
interfaces or traits define the signatures of the methods and
\emph{may} have some behaviour (such as in Java8 interfaces or Scala
traits). However classes are still responsible for object
instantiation and handling state. Prototype-based languages started
with Self~\cite{Ungar87self}, and are still widely used today via languages like
JavaScript. Frequentely prototype-based languages are dynamically 
typed. IB programming is clearly closer to CB programming. However the
key difference is that classes are not needed in IB
programming. Object interfaces recover the responsability of object
instantiation, and abstract state operations provide a way to deal
with state. A primary motivation of IB programming is to improve 
modularity and flexibility. Central to these goals are the 
more expressive and automated forms of type-refinement, and support 
for multiple interface inheritance.

@@

One famous attempt in this direction is provided by
\emph{MyType}~\cite{bruce1994paradigmatic}, representing the type of
\textbf{this}, changing its meaning along with inheritance. However when
invoking a method with MyType in a parameter position, the exact type of the
receiver must be known.  This is a big limitation in class based object oriented
programming, and is exasperated by the interface-based programming we propose: no
type is ever going to be exact since classes are not explicitly used. A recent
article~\cite{Saito2013933} lights up this topic, proposing two
new features: exact statements and nonheritable methods. Both are
related to our work: 1) any method generated inside of the \Q@of@ method is indeed
non-inheritable since there is no class name to extend from; 2) exact
statements (a form of wild-card capture on the exact run-time type) could
capture the ``exact type'' of an object even in a class-less
environment.
Admittedly, MyType greatly enhances the expressiveness and extensibility of
object-oriented programming languages. Object interfaces use covariant-return types to
simulate some uses of MyType. But this approach only works for refining
return types, whereas MyType is more general, as it also works for
parameter types. Our approach to covariantly refine state can recover
some of the additional expressiveness of MyType. As illustrated with our examples and
case studies, object interfaces are still very useful in many
practical applications, yet they do not require additional 
complexity from the type system.

>>>>

\paragraph{Models of Object-Oriented programming}
There are traditionally two big styles of OOP: class-based and
prototype-based. The class-based tradition started with Simula~\cite{Dahl66simula},
and is the dominant style of OOP today. Major languages such
as Java, C\#, C++ or Scala favour a class based OOP style. Very often
class-based languages are statically typed and they have interfaces or
some other similar mechanisms such as traits. However, in contrast to IB
programming, interfaces or traits in CB programming are not meant to
replace classes, but rather, they complement classes. Normally
interfaces or traits define the signatures of the methods and
\emph{may} have some behaviour (such as in Java 8 interfaces or Scala
traits). However classes are still responsible for object
instantiation and handling state. Prototype-based languages started
with Self~\cite{Ungar87self}, and are still widely used today via languages like
JavaScript. Frequently prototype-based languages are dynamically
typed. IB programming is clearly closer to CB programming. However the
key difference is that classes are not needed in IB
programming. Object interfaces recover the responsibility of object
instantiation, and abstract state operations provide a way to deal
with state. A primary motivation of IB programming is to improve
modularity and flexibility. Central to these goals are the
more expressive and automated forms of type-refinement, and support
for multiple interface inheritance.

@@

One famous attempt in this direction is provided by
\emph{MyType}~\cite{bruce1994paradigmatic}, representing the type of
\textbf{this}, changing its meaning along with inheritance. However when
invoking a method with MyType in a parameter position, the exact type of the
receiver must be known.  This is a big limitation in class based object oriented
programming, and is exasperated by the interface-based programming we propose: no
type is ever going to be exact since classes are not explicitly used. A recent
article~\cite{Saito2013933} lights up this topic, proposing two
new features: exact statements and nonheritable methods. Both are
related to our work: 1) any method generated inside the \Q@of@ method is indeed
non-inheritable since there is no class name to extend from; 2) exact
statements (a form of wild-card capture on the exact run-time type) could
capture the ``exact type'' of an object even in a class-less
environment.
Admittedly, MyType greatly enhances the expressiveness and extensibility of
object-oriented programming languages. Object interfaces use covariant-return types to
simulate some uses of MyType. But this approach only works for refining
return types, whereas MyType is more general, as it also works for
parameter types. Our approach to covariantly refine state can recover
some of the additional expressiveness of MyType. As illustrated with our examples and
case studies, object interfaces are still very useful in many
practical applications, yet they do not require additional
complexity from the type system.
